### 2019-09-04 experiments.tex
edit exp

In order to evaluate our approach and implementation, we applied
nondeterminism detection to real Python code, including some widely
used systems ({\tt redis-py} and {\tt pyfakefs} are well-maintained,
widely-used libraries).  The primary
points we wanted to explore were (1) whether our approach was able to reliably detect actual
nondeterminism, (2) whether the overhead of nondeterminism
detection for real systems was acceptable, and (3) the performance of
delta-debugging for real nondeterministic predicates.

\subsection{Case Study: Parallel Sorting}

Our first case study is a simple implementation of a parallel merge
sort in Python
\url{https://gist.github.com/stephenmcd/39ded69946155930c347}.  The
merge sort and harness can be found in the TSTL GitHub repository
\cite{tstl}, under {\tt examples/parallelsorts}.  When a sorting
algorithm is implemented using concurrency, a key question to ask is
whether the ordering of elements that are equal under the comparison
operator is consistent across sorts of the same original sequence.
This is related to the question of whether a sort is a stable sort
(preserving the original relative order of equal-by-comparison elements),
but imposes a less strict requirement on the sort; a concurrent sort
might not be stable, but might still produce the same (non-stable)
ordering from the same provided sequence, every time.  This is a
particular instance of a common expectation noted above:  while a parallel
implementation of an algorithm will likely have internal
nondeterminism (due to scheduling of work), it is often expected to
behave deterministically, from the point of view of a caller.

This case study has two purposes:  first, to show that the
nondeterminism detection features of TSTL can be used to, with
acceptable overhead, show that a parallel sort implementation provides
a consistent, deterministic ordering, and second to show that the
mechanism can also detect the nondeterminism in a parallel sort that
does exhibit nondeterministic behavior.  For the second purpose, we
implemented a very simple swap-based parallel sorting algorithm that
repeatedly has multiple threads scan through a sequence and swap
out-of-order neighbors, until the sequence is sorted (a kind of
parallel bubble-sort with a simpler termination criteria).

@@

\subsection {Case Study: {\tt redis-py}}

@@

\subsection{Case Study: {\tt datarray} Inference Algorithms}

@@

\subsection {Vertical Determinism Case Study: {\tt pyfakefs}}

@@

@@

Another threat to validity is posed by implementation errors.
TSTL and the mutation tool are tested for bugs by a set of Travis CI
tests, and are (for academic research code) moderately well-used by
external users, making this a lesser concern.  The nondeterminism
detected by TSTL can be demonstrated with standalone tests TSTL
generates that do not depend on any TSTL code.
The source code for the tools and all case studies is
available for inspection and replication in the TSTL repository.

>>>>

In order to evaluate our approach and implementation, we applied
nondeterminism detection to real Python code, including some widely
used systems ({\tt redis-py} and {\tt pyfakefs} are well-maintained,
widely-used libraries).  The primary
points we wanted to explore were (1) whether our approach was able to reliably detect actual
nondeterminism, (2) whether the overhead of nondeterminism
detection for real systems was acceptable, and (3) the performance of
delta-debugging for real nondeterministic predicates.  We chose two of
the subjects, {\tt redis-py} and {\tt pyfakefs}, based on the fact that they were previously-existing large
examples of TSTL harnesses, and therefore serve to show how easily our
approach can be integrated into an existing large-scale test
generation effort.  The other two subjects were chosen, and harnesses
written, specifically to demonstrate specific, important sources of
nondeterminism in Python code: concurrency and hash salting.

Table \ref{tab:overviewexp} shows a summary of our experimental
results.  In all cases, our approach was able to detect interesting
nondeterminism.  The overhead on testing imposed by nondeterminism
checks varied considerably, but in the worst case did not double the
average cost of each action, and thus never halved the amount of
testing performed.  Test reduction always reduced the size of
generated tests by 90\% or more, thus showing that effective reduction
has a large payoff.

\begin{table}
\centering
\begin{tabular}{l|l|r|r|l}
Subject & Detection & Overhead & Reduction & Reduction time \\
\hline
Parallel Sort & Yes (see below) & $\sim$40\% & & $\sim$2 minutes\\
{\tt redis-py} & Yes (see below) & $\sim$20\% & $\sim$95\% & $\sim$10 minutes\\
{\tt datarray} & Yes (see below) & $\sim$93\% & $\sim$90\% & $\sim$92
  seconds\\
{\tt pyfakefs} & Yes (see below) & $\sim$8\% & $\sim$99\% & $\sim$1 second  \\
\end{tabular}
\caption{Overview of experimental results}
\label{tab:overviewexp}
\end{table}

\subsection{Parallel Sorting}

Our first subject is a simple implementation of a parallel merge
sort\footnote{\url{https://gist.github.com/stephenmcd/39ded69946155930c347}}.  The
merge sort and harness can be found in the TSTL GitHub repository
\cite{tstl}, under {\tt examples/parallelsorts}.  When a sorting
algorithm is implemented using concurrency, a key question to ask is
whether the ordering of elements that are equal under the comparison
operator is consistent across sorts of the same original sequence.
This is related to the question of whether a sort is a stable sort
(preserving the original relative order of equal-by-comparison elements),
but imposes a less strict requirement on the sort; a concurrent sort
might not be stable, but might still produce the same (non-stable)
ordering from the same provided sequence, every time.  This is a
particular instance of a common expectation noted above:  while a parallel
implementation of an algorithm will likely have internal
nondeterminism (due to scheduling of work), it is often expected to
behave deterministically, from the point of view of a caller.

This experimental subject serves two purposes:  first, to show that the
nondeterminism detection features of TSTL can be used to, with
acceptable overhead, show that a parallel sort implementation provides
a consistent, deterministic ordering, and second to show that the
mechanism can also detect the nondeterminism in a parallel sort that
does exhibit nondeterministic behavior.  For the second purpose, we
implemented a very simple swap-based parallel sorting algorithm that
repeatedly has multiple threads scan through a sequence and swap
out-of-order neighbors, until the sequence is sorted (a kind of
parallel bubble-sort with a simpler termination criteria).

@@

\subsection {The {\tt redis-py} Library}

@@

\subsection{Berkeley {\tt datarray} Inference Algorithms}

@@

\subsection {Vertical Determinism: {\tt pyfakefs}}

@@

\subsection{Summary of Results}

@@

Another threat to validity is posed by implementation errors.
TSTL and the mutation tool are tested for bugs by a set of Travis CI
tests, and are (for academic research code) moderately well-used by
external users, making this a lesser concern.  The nondeterminism
detected by TSTL can be demonstrated with standalone tests TSTL
generates that do not depend on any TSTL code.
The source code for the tools and all experimental subjects is
available for inspection and replication in the TSTL repository.
